---
title: "Thesis Cleaning"
output: html_document
date: "2023-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)    
library(tidyr)
library(broom)
library(dslabs)
ds_theme_set()
library(ggplot2)
library(readr)
library(knitr)
library(gridExtra)
library(dplyr)
library(table1)
library(splitstackshape)
library(caret)
library(e1071)
library(pROC)
library(randomForest)
library(rpart)
library(rpart.plot)
library(kableExtra)
library(scales)
library(readxl)
```

## Loading in data

```{r}
AZ <- read_xlsx("/Users/nicolekaufmann/Desktop/Thesis/AZ COVID-19 Data (FINAL deidentified) wave 1 (1).xlsx")
AZ <- AZ[-1,]
AZ 
CA <- read_xlsx("/Users/nicolekaufmann/Desktop/Thesis/CA COVID-19 Data (FINAL deidentified) wave 1.xlsx")
CA <- CA[-1,]
CA
str(AZ)
table(AZ$Q7)

```


```{r}
#Cleaning AZ
library(dplyr)
#Dropping if answer "I don't know/Prefer not to answer" to Q14 on the AZ survey (willingness to get COVID vaccine)
AZ <- AZ %>% filter(Q14 != "5")
#Renaming Q14= "Vax_Likelihood"
AZ <- AZ %>% rename("Vax_Likelihood"="Q14")
AZ
#Cleaning CA
#Dropping if answer "I don't know/Prefer not to answer" to Q6 on the CA survey
#First, renaming "6" to be "Vax Likelihood"
CA <- CA%>% rename("Vax_Likelihood"="6")
#Now, filtering out people who responded "I don't know" to outcome variable
CA <- CA %>% filter(Vax_Likelihood != "I don't know/prefer not to answer") %>% filter(Vax_Likelihood != "No lo sé / prefiero no responder")

#Renaming race column
CA <- CA %>% rename("Race_Question"="1...7")
str(CA)
table(CA$Race_Question)

#Renaming urban/suburban location question for AZ and CA
CA <- CA%>% rename("geo_location"="49")
AZ <- AZ%>% rename("geo_location"="Q84")

```

```{r}
#Changing Sí to Yes for all of CA data 
CA[CA=="Sí"] <- "Yes"

#Translating "No lo sé / prefiero no responder" to "I don't know/prefer not to answer" for all of CA data
CA[CA=="No lo sé / prefiero no responder"] <- "I don't know/prefer not to answer"

```


Now, I will create race variables
```{r}
#For AZ, race variable is Q7: "What race/ethnicity do you most identify with? Select all that apply." We will create two categories: 1) Hispanic/Latino and 2) multi-ethnic (1 is Hispanic/Latino, 1+ anything else is multi-ethnic)
table(AZ$Q7, useNA = "always")
AZ <- AZ %>% mutate(Race_Question = ifelse(Q7=="1", "Hispanic/Latino", "Multi-Ethnic"))
table(AZ$Race_Question)
#Only 21 identify as multi-ethnic. NOT FILTERING THEM OUT JUST YET
#AZ <- AZ %>% filter(Race_Question =="Hispanic/Latino")
#table(AZ$Race_Question)

#FOR CA, I renamed it "Race_Question." It is a single multiple choice question: "What race/ethnicity do you most identify with?"
table(CA$Race_Question)
#First, renaming Spanish answers to English
CA[CA=="Hispano/latino (incluyendo herencia indígena)"] <- "Hispanic/Latino (including of Indigenous heritage)"
table(CA$Race_Question)
#For CA, now dropping all the people who do not identify as Hispanic/Latino
CA <- CA %>% filter(Race_Question == "Hispanic/Latino (including of Indigenous heritage)")
table(CA$Race_Question)
#Now, we only have people who identify as Hispanic/Latino and Multi-Ethnic

```

Now, I will create a urban/rural variable for the datasets
```{r}
#Dropping people who do not know/prefer not to answer (answered 4 to Q49)
CA[CA=="Urbana / ciudad"] <- "Urban/City"
CA[CA=="Suburbana"] <- "Suburban"
CA[CA=="Rural"] <- "Rural/County area"
CA[CA=="No sé / prefiero no responder"] <- "I don't know/prefer not to answer"

#Creating variable for CA dataset
CA <- CA %>% mutate(location = ifelse(geo_location == "Urban/City", 2,
                                      ifelse(geo_location == "Suburban", 1,
                                             ifelse(geo_location == "Rural/County area", 0, NA))))
table(CA$location, useNA = "always") #The NAs represent "I don't knows" or blanks
CA <- CA %>% filter(location != "NA")
table(CA$location, useNA = "always")

#Creating variable for AZ dataset
AZ <- AZ %>% mutate(location=ifelse(geo_location=="3",2,
                                    ifelse(geo_location=="2",1,
                                           ifelse(geo_location=="1",0, NA))))
table(AZ$location,useNA="always")
AZ <- AZ %>% filter(location != "NA")
table(AZ$location, useNA="always")
```

```{r}
#Making a Likelihood Variable. For AZ, if Vax_Likelihood=1 or 2, then "Likely" and if Vax_Likelihood=3 or 4, then "Unlikely." Likely=1 and unlikely=0. 
AZ <- AZ %>% mutate(Likely = ifelse(Vax_Likelihood==1, 1,
                                    ifelse(Vax_Likelihood==2, 1,
                                           ifelse(Vax_Likelihood==3,0,
                                                  ifelse(Vax_Likelihood==4,0, NA)))))

table(AZ$Likely, useNA = "always")

#Making a Likelihood Variable. For CA, if Extremely Likely or Extremadamente probable, then Likely =1. If Somewhat Likely or Algo probable, then Likely=1. If Unlikely or Poco Probable, then Likely=0. If Very Unlikely or Muy Improbable, then Likely = 0. 

CA <- CA %>% mutate(Likely = ifelse(Vax_Likelihood== "Extremely likely", 1,
                                    ifelse(Vax_Likelihood=="Extremadamente probable", 1,
                                           ifelse(Vax_Likelihood=="Algo probable",1,
                                                  ifelse(Vax_Likelihood=="Somewhat likely",1,
                                                         ifelse(Vax_Likelihood=="Unlikely",0,
                                                                ifelse(Vax_Likelihood=="Poco probable",0,
                                                                       ifelse(Vax_Likelihood=="Very unlikely", 0,
                                                                              ifelse(Vax_Likelihood=="Muy improbable",0, NA)))))))))

#Checking NAs
table(AZ$Likely, useNA = "always")
table(CA$Likely, useNA="always")
#No NAs exist for the "Likely" column

```

#For CA dataset, creating a variable for having at least one comorbid condition
```{r}
#First, we rename the conditions for both datasets
CA <- CA%>% rename("Heart_Issue"="17_1") %>% rename("Lung_Issue"="17_2") %>% rename("Cancer"="17_3") %>% rename("Hypertension"="17_4") %>% rename("Diabetes"="17_5") %>% rename("Obesity"="17_6") %>% rename("Autoimmune_Disorder"="17_7") %>% rename("Other_Autoimmune"="17_8") %>% rename("Depressive_Disorder"="17_9") %>% rename("Anxiety_Disorder"="17_10") %>% rename("Dementia_Disorder"="17_11")


AZ <- AZ %>% rename("Heart_Issue"="Q48_1") %>% rename("Lung_Issue"="Q48_2") %>% rename("Cancer"="Q48_3") %>% rename("Hypertension"="Q48_4") %>% rename("Diabetes"="Q48_5") %>% rename("Obesity"="Q48_6") %>% rename("Autoimmune_Disorder"="Q48_7") %>% rename("Other_Autoimmune"="Q48_8") %>% rename("Depressive_Disorder"="Q48_9") %>% rename("Anxiety_Disorder"="Q48_10") %>% rename("Dementia_Disorder"="Q48_11")

```
Not all of the above conditions are comorbid conditions. Anxiety and Depressive Disorders should not be considered comorbid conditions. 

Now, dropping people who do not have any comorbid conditions
```{r}
#If "I don't know/prefer not to answer" for all of the comorbid conditions, we should drop them. Creating variable if "I don't know/prefer not to answer" for all conditions

#Cleaning CA
CA$no_comorbid = (CA$Heart_Issue == "I don't know/prefer not to answer") + (CA$Lung_Issue == "I don't know/prefer not to answer") + (CA$Cancer == "I don't know/prefer not to answer") + (CA$Hypertension == "I don't know/prefer not to answer") + (CA$Diabetes == "I don't know/prefer not to answer") + (CA$Obesity == "I don't know/prefer not to answer") + (CA$Autoimmune_Disorder == "I don't know/prefer not to answer") + (CA$Other_Autoimmune == "I don't know/prefer not to answer") + (CA$Dementia_Disorder == "I don't know/prefer not to answer")
table(CA$no_comorbid)
#no one to kick out for CA
  
#Cleaning AZ
AZ$no_comorbid = (AZ$Heart_Issue ==3) + (AZ$Lung_Issue==3)+ (AZ$Cancer==3) + (AZ$Hypertension==3) + (AZ$Diabetes ==3) + (AZ$Obesity ==3) + (AZ$Autoimmune_Disorder==3) + (AZ$Other_Autoimmune==3) + (AZ$Dementia_Disorder==3) 

table(AZ$no_comorbid) #2 people answered "I don't know" to all - will filter them out. There were 4 when I looked in raw data, and 2 had not answered the urban question, so makes sense!
AZ <- AZ %>% filter(no_comorbid !=9)
```


```{r}
#For CA dataset: Creating a variable for having at least one comorbid condition. Comorbid=1 means has a comorbid condition. Comorbid=0 means the person does not have a  comorbid condition. 
CA <- CA %>% mutate(comorbidity = ifelse(Heart_Issue == "Yes", 1,
                                         ifelse(Lung_Issue =="Yes", 1,
                                          ifelse(Cancer == "Yes",1,
                                          ifelse(Hypertension == "Yes",1,
                                           ifelse(Diabetes == "Yes",1,
                                           ifelse(Obesity == "Yes",1,
                                ifelse(Autoimmune_Disorder == "Yes",1,
                             ifelse(Other_Autoimmune =="Yes",1, ifelse(Dementia_Disorder == "Yes",1, 0))))))))))
table(CA$comorbidity, useNA = "always")

#For AZ dataset: Creating a variable for having at least one comorbid condition. Comorbid=1 means has a comorbid condition. Comorbid=0 means the person does not have a  comorbid condition. 
AZ <- AZ %>% mutate(comorbidity = ifelse(Heart_Issue == "1", 1,
                                         ifelse(Lung_Issue =="1", 1,
                                          ifelse(Cancer == "1",1,
                                          ifelse(Hypertension == "1",1,
                                    ifelse(Diabetes == "1",1,
                                  ifelse(Obesity == "1",1,
                                ifelse(Autoimmune_Disorder == "1",1,
                             ifelse(Other_Autoimmune =="1",1, ifelse(Dementia_Disorder == "1",1, 0))))))))))
table(AZ$comorbidity, useNA = "always")
```

Now, creating variables for educational level
```{r}

#For AZ, Q82 is "what is your highest level of completed education"
#Grouping together all people who have never been to school, less than primary/elementary, and less than secondary/high school as "high school or less." There are only ~6 people who have never been to school/have less than elementary - too small of a stratum.

#Less than or equal to high school=0, college degree=1, graduate degree=2, "I don't know" is set to NA
AZ <- AZ %>% mutate(education=ifelse(Q82=="1", 0,
                               ifelse(Q82=="2",0,
                               ifelse(Q82=="3",0,
                              ifelse(Q82=="4",1,
                              ifelse(Q82=="5",2, NA))))))
table(AZ$education, useNA="always")
#Dropping people who responded "6," which was "I don't know/prefer not to answer"
AZ <- AZ %>% filter(education !="6")
table(AZ$education, useNA="always")

#For CA survey, education question is Q40
table(CA$"40")

CA <- CA %>% rename("ed_level"="40") 
table(CA$ed_level, useNA = "always")
CA <-CA %>% mutate(education= ifelse(ed_level=="Escuela primaria o menos",0,
                              ifelse(ed_level=="Primary/elementary school or less",0,
                                ifelse(ed_level=="Escuela secundaria",0,
                              ifelse(ed_level=="Secondary/high school",0,
                                     ifelse(ed_level=="Título del colegio comunitario o universitario",1,
                                  ifelse(ed_level=="College or University degree",1,
                              ifelse(ed_level=="Graduate/Postgraduate degree",2,
                                     ifelse(ed_level=="Título graduado o posgrado",2, NA)))))))))
table(CA$education, useNA="always")
#Dropping people who did not respond or who responded "I don't know/prefer not to answer"
CA <- CA %>% filter(education != "NA")
table(CA$education, useNA = "always")

```

Now, I will create a variable for gender for CA
```{r}
#For CA, asked only "how would you describe your sex" - male, female, other, prefer not to answer in Q36.
#mujer = female, hombre=man

CA[CA=="Mujer"] <- "Female"
CA[CA=="Hombre"] <- "Male"

#if gender=1, male. if gender=0, female. 
CA <- CA %>% rename(self_sex = "36") %>% mutate(gender = ifelse(self_sex=="Male", 1,
                                                      ifelse(self_sex=="Female",0, NA)))

table(CA$self_sex)
table(CA$gender, useNA = "always")
#Filtering out people who did not answer

CA <- CA %>% filter(gender != "NA")
table(CA$gender, useNA = "always")
```

Now, I will create gender variable for AZ
```{r}
#For AZ, Q76 is "how would you describe yourself?"
table(AZ$"Q76", useNA="always")
AZ <- AZ %>% rename(self_sex="Q76") %>% mutate(gender= ifelse(self_sex=="1",1,
                                                           ifelse(self_sex=="2",0, NA)))

table(AZ$self_sex, useNA = "always")
table(AZ$gender, useNA="always")
#filtering out people who identify as neither male nor female
AZ <- AZ %>% filter(gender != "NA")
table(AZ$gender)

```

Now, I will make variables for PHQ4 scores for Arizona
```{r}
#For AZ, Q52 is the PHQ4 question. I am creating a variable that is =1 if the participant answered "5" or "I prefer not to answer" to at least one of the questions. This is because the PHQ4 question relies on data from every column - cannot accurately detect anxiety/depression otherwise. 

#If PHQ_empty=1, then do not use
AZ <- AZ %>% mutate(PHQ_empty = ifelse(Q52_1=="5", 1,
                                 ifelse(Q52_2=="5", 1,
                                 ifelse(Q52_3=="5",1,
                               ifelse(Q52_4=="5",1,0)))))

table(AZ$PHQ_empty)
#Filtering out people for whom there is no PHQ4 score (PHQ empty =1)
AZ <- AZ %>% filter(PHQ_empty == "0")
table(AZ$PHQ_empty, useNA="always")

#Creating depression score. This means Q52_1 + Q52_2 >=3 is depressed
AZ$Q52_1 <- as.numeric(AZ$Q52_1)
AZ$Q52_2 <- as.numeric(AZ$Q52_2)
AZ$Q52_3 <- as.numeric(AZ$Q52_3)
AZ$Q52_4 <- as.numeric(AZ$Q52_4)

#If depressed=1, if not=0
#filter out invalid PHQ
AZ <- AZ %>% filter(PHQ_empty != "1")
#First have to change the scale of the questions
AZ <- AZ %>% mutate(little_interest = ifelse(Q52_1=="1", 3,
                                          ifelse(Q52_1=="2", 2,
                                        ifelse(Q52_1=="3",1, 0))))

AZ <- AZ %>% mutate(feeling_down = ifelse(Q52_2=="1", 3,
                                          ifelse(Q52_2=="2", 2,
                                        ifelse(Q52_2=="3",1, 0))))
                                              
AZ <- AZ %>% mutate(sum_dep = (little_interest + feeling_down)) %>% mutate(PHQ_dep = ifelse(sum_dep >= 3, 1,0))
table(AZ$sum_dep)
#If depressed, PHQ_dep=1
table(AZ$PHQ_dep)

#Now, will repeat for anxiety
AZ <- AZ %>% mutate(nervous = ifelse(Q52_3=="1", 3,
                                          ifelse(Q52_3=="2", 2,
                                        ifelse(Q52_3=="3",1, 0))))

AZ <- AZ %>% mutate(worrying = ifelse(Q52_4=="1", 3,
                                          ifelse(Q52_4=="2", 2,
                                        ifelse(Q52_4=="3",1, 0))))

AZ <- AZ %>% mutate(sum_anx = (nervous + worrying)) %>% mutate(PHQ_anx = ifelse(sum_anx >=3, 1,0))
table(AZ$PHQ_anx)

```

Now, I will make PHQ4 scores for CA. There is no option for "I don't know/prefer not to answer" for CA.
```{r}
#Double-Check translations
CA <- CA %>% rename("PH_int"="19_1") %>% mutate(little_interest = ifelse(PH_int=="Nearly every day", 3,
                                      ifelse(PH_int=="Casi a diario", 3,       
                                             ifelse(PH_int=="More than half the days",2,
                                                    ifelse(PH_int=="La mayoría de los días",2,
                                             ifelse(PH_int == "Several days", 1,
                                                   ifelse(PH_int == "Varios días", 1, 
                                              ifelse(PH_int=="Not at all",0,
                                                     ifelse(PH_int=="Ningún día",0,NA)))))))))
table(CA$little_interest)
CA <- CA %>% rename("PH_feel"="19_2") %>% mutate(feeling_down = ifelse(PH_feel=="Nearly every day", 3,
                                      ifelse(PH_feel=="Casi a diario", 3,       
                                             ifelse(PH_feel=="More than half the days",2,
                                                    ifelse(PH_feel=="La mayoría de los días",2,
                                             ifelse(PH_feel == "Several days", 1,
                                                   ifelse(PH_feel == "Varios días", 1, 
                                              ifelse(PH_feel=="Not at all",0,
                                                     ifelse(PH_feel=="Ningún día",0,NA)))))))))
table(CA$feeling_down)

#Making depression variable
CA <- CA%>% mutate(sum_dep = feeling_down + little_interest) %>% mutate(PHQ_dep= ifelse(sum_dep >=3, 1, 0))
table(CA$PHQ_dep)

CA <- CA %>% rename("PH_nervous"="19_3") %>% mutate(nervous = ifelse(PH_nervous=="Nearly every day", 3,
                                      ifelse(PH_nervous=="Casi a diario", 3,       
                                             ifelse(PH_nervous=="More than half the days",2,
                                                    ifelse(PH_nervous=="La mayoría de los días",2,
                                             ifelse(PH_nervous == "Several days", 1,
                                                   ifelse(PH_nervous == "Varios días", 1, 
                                              ifelse(PH_nervous=="Not at all",0,
                                                     ifelse(PH_nervous=="Ningún día",0,NA)))))))))
table(CA$nervous)

CA <- CA %>% rename("PH_worrying"="19_4") %>% mutate(worrying = ifelse(PH_worrying=="Nearly every day", 3,
                                      ifelse(PH_worrying=="Casi a diario", 3,       
                                             ifelse(PH_worrying=="More than half the days",2,
                                                    ifelse(PH_worrying=="La mayoría de los días",2,
                                             ifelse(PH_worrying == "Several days", 1,
                                                   ifelse(PH_worrying == "Varios días", 1, 
                                              ifelse(PH_worrying=="Not at all",0,
                                                     ifelse(PH_worrying=="Ningún día",0,NA)))))))))
table(CA$worrying)

#Making anxiety variable
CA <- CA %>% mutate(sum_anx= nervous + worrying) %>% mutate(PHQ_anx = ifelse(sum_anx >=3, 1,0))

```

Checking age data for CA
```{r}
#Question 37 on CA questionnaire
CA[CA=="32 anos"] <- "32"
table(CA$"37", useNA="always")
#Renaming column
CA <- CA %>% rename(Age="37")
class("Age")

#It is a character, so changing it to numeric
CA <- CA %>% mutate(Age= as.numeric(Age))
table(CA$Age, useNA="always")

#checking age distribution for CA
age_dist <-  ggplot(CA, aes(x=Age))+
  geom_histogram(binwidth=1, color="blue", fill="white")+
  xlab("Age")+
  ylab("Count")+
  ggtitle("Age Distribution in CA")

age_dist

#Now checking for AZ
table(AZ$"Q79")
str(AZ)
view(AZ)
#Q79 in AZ questionnaire
AZ <- AZ %>% rename("Age"="Q79")
table(AZ$Age, useNA = "always")
#Filtering out NA, changing class
AZ <- AZ %>% filter(Age !="NA")
class(AZ$Age)
#Also a character variable, so changing to numeric
AZ <- AZ %>% mutate(Age= as.numeric(Age))
class(AZ$Age)

#checking age distribution for AZ
age_dist1 <-  ggplot(AZ, aes(x=Age))+
  geom_histogram(binwidth=1, color="blue", fill="white")+
  xlab("Age")+
  ylab("Count")+
  ggtitle("Age Distribution in AZ")

age_dist1
```

```{r}
#Create new data set
AZ_new <- AZ %>% select(Vax_Likelihood, Likely, location, Race_Question, comorbidity, education, gender, PHQ_dep,PHQ_anx, Age)
CA_new <- CA %>% select(Vax_Likelihood, Likely, location, Race_Question, comorbidity, education, gender, PHQ_dep,PHQ_anx, Age)

combined_data <- full_join(AZ_new, CA_new)
view(combined_data)

```








